---
title: "Using RUMBLE for Microbiome Biomarker Discovery"
author: "Lucas Fett"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Using RUMBLE for Microbiome Biomarker Discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 7,
  fig.align = "center",
  out.width = "100%",
  dpi = 72,
  warning = FALSE,
  message = FALSE,
  crop = NULL
)
```

# Introduction

The **RUMBLE** (Robust Unified Microbiome Biomarker Learning Engine) package provides a streamlined workflow for identifying robust microbial biomarkers from high-throughput sequencing data using a consensus machine learning strategy.

Version 1.1.0 integrates classical Differential Abundance (DA) analysis alongside model-agnostic interpretability (SHAP). This allows researchers to validate complex predictive patterns with rigorous False Discovery Rate (FDR) control using methods such as ALDEx2, ANCOM-BC, or Wilcoxon.

This vignette demonstrates a simplified exploratory workflow using the example phyloseq object distributed with the package.

# Load Example Data

```{r load_data}
library(RUMBLE)
library(phyloseq)
library(knitr)

ps_path <- system.file(
  "extdata",
  "PRJEB38465_phyloseq_com_metadados_completos.rds",
  package = "RUMBLE"
)

if (ps_path == "") {
  stop("Example data not found. Check package installation.")
}

ps <- readRDS(ps_path)
print(ps)
```

# Run a Fast Exploratory Analysis

To keep the vignette lightweight, we reduce `grid_size` and `shap_reps`. In real analyses, prefer default values (`grid_size = 30`, `shap_reps = 25`).

```{r run_pipeline}
set.seed(42)

results_fast <- RUMBLE(
  input             = ps,
  outcome_var       = "ses",
  class_of_interest = "Low",
  tax_level         = "Genus",
  project_name      = "Fast_Exploration",
  output_dir        = tempdir(),
  grid_size         = 10,
  shap_reps         = 2,
  xgb_trees         = 100,
  rf_trees          = 100,
  n_cores           = 1,
  shap_data         = "test",
  run_da            = TRUE,
  da_method         = "wilcoxon"
)
```

# Performance Summary

```{r metrics_table}
knitr::kable(
  results_fast$metrics,
  digits = 3,
  caption = "Model performance metrics for the exploratory analysis."
)
```

# Global Results

## Metrics Comparison

```{r metrics_plot}
print(results_fast$plots$metrics)
```

## ROC Curves

```{r roc_plot}
print(results_fast$plots$roc)
```

## Confusion Matrices

```{r cm_plot}
print(results_fast$plots$cm)
```

## SHAP Consensus (Spearman)

```{r shap_spearman}
print(results_fast$plots$shap_spearman)
```

## Global SHAP Beeswarm

```{r shap_beeswarm}
print(results_fast$plots$shap_beeswarm)
```

## Taxa Prevalence

```{r shap_prevalence}
print(results_fast$plots$shap_prevalence)
```

## Integrated Biomarker Interpretation (Spearman)

```{r biomarker_integrated, fig.height=9, fig.width=12, fig.cap="Combined view of SHAP importance (Spearman) and group prevalence."}
print(results_fast$plots$biomarker_integrated_spearman)
```

# Exporting Data for Custom Analysis

The RUMBLE pipeline automatically generates an integrated data frame that mirrors the biomarker dashboard.

```{r integrated_table}
integrated_df <- results_fast$integrated_table
head(integrated_df)
```

The table includes:

- **Taxon**: Feature name  
- **SHAP_Score**: Mean absolute SHAP value  
- **Direction_Class**: Associated class  
- **Prevalence_***: Presence across groups  
- **logFC, p_val, adj_p_val**: Differential abundance statistics  

Additionally, two `.tsv` files are generated in `output_dir`:

- `[project]_integrated_biomarkers.tsv`  
- `[project]_top_differential_abundance.tsv`  

# Permutation Importance

```{r perm_heat}
print(results_fast$plots$perm_heat)
```

# Global Importance Tables

```{r importance_table}
knitr::kable(
  head(results_fast$importance$global_importance$spearman, 10),
  digits = 4,
  caption = "Top 10 biomarkers by SHAP consensus (Spearman direction)."
)
```

# Model-Specific Outputs

## SHAP Beeswarm per Model

```{r model_beeswarm}
model_names <- names(results_fast$plots$model_specific$shap_beeswarm)

for (model_name in model_names) {
  cat("### ", model_name, " beeswarm\n\n", sep = "")
  print(results_fast$plots$model_specific$shap_beeswarm[[model_name]])
  cat("\n")
}
```

## Integrated Biomarker Interpretation per Model

```{r model_integrated, fig.height=9, fig.width=12}
model_names <- names(results_fast$plots$model_specific$biomarker_integrated_spearman)

for (model_name in model_names) {
  cat("### ", model_name, " integrated biomarker interpretation (Spearman)\n\n", sep = "")
  print(results_fast$plots$model_specific$biomarker_integrated_spearman[[model_name]])
  cat("\n")
}
```

# Session Information

```{r session_info}
sessionInfo()
```
