---
title: "Using RUMBLE for Training-Set Interpretability and Test-Set Performance Assessment"
author: "Lucas Fett"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Using RUMBLE for Training-Set Interpretability and Test-Set Performance Assessment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 9,
    fig.height = 7,
    fig.align = "center",
    out.width = "100%",
    dpi = 72,
    warning = FALSE,
    message = FALSE,
    crop = NULL
)
```

# Introduction

The **RUMBLE** (Robust Unified Microbiome Biomarker Learning Engine)
package provides an end-to-end workflow for microbiome biomarker discovery
using consensus machine learning, SHAP-based interpretability, permutation
importance, and the TreeSHAP S-R-I ecological decomposition.

This vignette is organized around two complementary analyses using the
package example dataset.

In the **first analysis**, the full interpretability workflow is performed on
the **training set**, using **CLR normalization** and **2 SHAP repetitions**.
This section shows the complete interpretability output, including SHAP,
permutation importance, prevalence plots, integrated biomarker plots, and the
TreeSHAP ecology layer.

In the **second analysis**, the pipeline is rerun with SHAP evaluated on the
**test set**, but the vignette displays only the **model performance table**
and the **performance-oriented plots**. This creates a clear separation
between model interpretation during training and external performance
assessment on held-out data.

# Load Packages and Example Data

```{r load_data}
library(RUMBLE)
library(phyloseq)
library(dplyr)
library(ggplot2)

ps_path <- system.file(
    "extdata",
    "PRJEB38465_phyloseq_com_metadados_completos.rds",
    package = "RUMBLE"
)

if (ps_path == "") {
    stop("Example data not found. Check package installation.")
}

ps <- readRDS(ps_path)
print(ps)
```

## Metadata Overview

```{r metadata_overview}
meta <- as.data.frame(sample_data(ps))
cat("Target variable distribution (ses):\n")
print(table(meta$ses))
```

# Analysis 1: Full Training-Set Interpretability with CLR and 2 SHAP Repetitions

This first run is the main interpretability analysis presented in the
vignette. It uses the **training set** for SHAP evaluation, applies standard
**CLR normalization**, and sets **`shap_reps = 2`**.

To keep the vignette computationally realistic, the tree counts and tuning
grid remain moderate, but the interpretability stage follows the requested
configuration.

```{r run_train_analysis, results='hide'}
set.seed(42)

results_train <- RUMBLE(
    input                = ps,
    outcome_var          = "ses",
    class_of_interest    = "Low",
    tax_level            = "Genus",
    project_name         = "Vignette_Train_CLR",
    output_dir           = tempdir(),
    cv_folds             = 2,       
    grid_size            = 2,       
    shap_reps            = 10,      
    xgb_trees            = 50,      
    rf_trees             = 50,      
    n_cores              = 1,
    shap_data            = "train",
    normalization_method = "clr",
    shap_method          = "fast",  
    verbose              = FALSE      
)
```

## Performance Table

```{r train_metrics_table}
knitr::kable(
    results_train$metrics,
    digits = 3,
    caption = "Model performance metrics for the training-set interpretability run."
)
```

## Performance Plots

### Metrics Comparison

```{r train_metrics_plot, fig.height=5, fig.cap="Comparison of model performance metrics for the training-set interpretability run."}
print(results_train$plots$metrics)
```

### ROC Curves

```{r train_roc_plot, fig.height=5, fig.cap="ROC curves for the training-set interpretability run."}
print(results_train$plots$roc)
```

### Confusion Matrices

```{r train_cm_plot, fig.height=8, fig.cap="Confusion matrices for the training-set interpretability run."}
print(results_train$plots$cm)
```

## SHAP Consensus

RUMBLE summarizes the global SHAP signal using two complementary direction
criteria: **Spearman correlation** between feature abundance and contribution,
and **mean SHAP sign** across observations.

```{r train_shap_spearman, fig.height=8, fig.cap="Top biomarkers ranked by SHAP consensus using Spearman direction on the training set."}
print(results_train$plots$shap_spearman)
```

```{r train_shap_mean, fig.height=8, fig.cap="Top biomarkers ranked by SHAP consensus using mean SHAP direction on the training set."}
print(results_train$plots$shap_mean)
```

## SHAP Beeswarm

```{r train_beeswarm, fig.height=8, fig.cap="Beeswarm plot of SHAP values on the training set."}
print(results_train$plots$shap_beeswarm)
```

## Taxa Prevalence

```{r train_prevalence, fig.height=7, fig.cap="Prevalence of top SHAP features across outcome groups on the training set."}
print(results_train$plots$shap_prevalence)
```

## Integrated Biomarker Interpretation

```{r train_integrated_mean, fig.height=9, fig.width=12, fig.cap="Integrated interpretation combining mean SHAP importance and prevalence on the training set."}
print(results_train$plots$biomarker_integrated_mean)
```

```{r train_integrated_spearman, fig.height=9, fig.width=12, fig.cap="Integrated interpretation combining Spearman-based SHAP importance and prevalence on the training set."}
print(results_train$plots$biomarker_integrated_spearman)
```

## Permutation Importance

```{r train_perm_heat, fig.height=8, fig.cap="Permutation importance heatmap for the training-set interpretability run."}
print(results_train$plots$perm_heat)
```

## TreeSHAP S-R-I Ecology Layer

```{r train_sri_decomposition, fig.height=8, fig.cap="Stacked S-R-I decomposition of the top taxa for the training-set interpretability run."}
print(results_train$plots$sri_decomposition)
```

```{r train_sri_profile, fig.height=8, fig.cap="Ecological profile of the top taxa derived from the TreeSHAP S-R-I layer."}
print(results_train$plots$sri_ecological_profile)
```

## Global Importance Tables

```{r train_importance_spearman}
knitr::kable(
    head(results_train$importance$global_importance$spearman, 10),
    digits = 4,
    caption = "Top 10 biomarkers by SHAP consensus using Spearman direction on the training set."
)
```

```{r train_importance_mean}
knitr::kable(
    head(results_train$importance$global_importance$mean_shap, 10),
    digits = 4,
    caption = "Top 10 biomarkers by SHAP consensus using mean SHAP direction on the training set."
)
```

## Tree Ecology Tables

```{r train_sri_profile_table}
knitr::kable(
    head(results_train$tree_ecology$profile, 10),
    digits = 4,
    caption = "Top taxa and their S-R-I ecological decomposition profile."
)
```

```{r train_synergy_edges_table}
knitr::kable(
    head(results_train$tree_ecology$synergy_edges, 10),
    digits = 4,
    caption = "Top synergy edges inferred from the TreeSHAP interaction layer."
)
```

```{r train_redundancy_edges_table}
knitr::kable(
    head(results_train$tree_ecology$redundancy_edges, 10),
    digits = 4,
    caption = "Top redundancy edges inferred from the TreeSHAP interaction layer."
)
```

