---
title: "Using RUMBLE for Microbiome Biomarker Discovery"
author: "Lucas Fett"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Using RUMBLE for Microbiome Biomarker Discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 9,
    fig.height = 7,
    fig.align = "center",
    out.width = "100%",
    dpi = 72,
    warning = FALSE,
    message = FALSE,
    crop = NULL
)
```

# Introduction

The **RUMBLE** (Robust Unified Microbiome Biomarker Learning Engine)
package provides a streamlined workflow for identifying robust microbial
biomarkers from high-throughput sequencing data using a consensus machine
learning strategy.

This simplified vignette focuses on a single exploratory workflow using the
example `phyloseq` object distributed with the package. The goal is to show a
fast run of the full pipeline and then inspect the main outputs returned by
`RUMBLE()`, including both consensus-level and model-specific interpretability
plots.

# Load example data

We begin by loading the package and the example dataset.

```{r load_phyloseq}
library(RUMBLE)
library(phyloseq)
library(knitr)

ps_path <- system.file(
    "extdata",
    "PRJEB38465_phyloseq_com_metadados_completos.rds",
    package = "RUMBLE"
)

if (ps_path == "") {
    stop("Example data not found. Check package installation.")
}

ps <- readRDS(ps_path)
print(ps)
```

# Run a fast exploratory analysis

The following chunk runs the exploratory workflow with reduced computational
settings so that the vignette remains lightweight while still demonstrating the
complete analysis pipeline.

```{r run_exploratory, results='hide'}
set.seed(42)

# Run pipeline with reduced parameters for speed
results_fast <- RUMBLE(
    input             = ps,
    outcome_var       = "ses",
    class_of_interest = "Low",        
    tax_level         = "Genus",
    project_name      = "Fast_Exploration",
    output_dir        = tempdir(),
    grid_size         = 10,
    shap_reps         = 2,
    xgb_trees         = 100,
    rf_trees          = 100,
    n_cores           = 4,
    shap_data         = "test" # Using test set for SHAP
)
```

# Performance summary

The metrics table provides a compact overview of predictive performance for all
trained models.

```{r metrics_table_ps}
knitr::kable(
    results_fast$metrics,
    digits = 3,
    caption = "Model performance metrics for the exploratory analysis."
)
```

# Global results

The sections below display the main global outputs produced by the pipeline.

## Metrics comparison

```{r plot_ps_metrics, fig.height=5, fig.cap="Performance comparison across fitted models."}
print(results_fast$plots$metrics)
```

## ROC curves

```{r plot_ps_roc, fig.height=5, fig.cap="ROC curves for all fitted models."}
print(results_fast$plots$roc)
```

## Confusion matrices

```{r plot_ps_cm, fig.height=8, fig.cap="Confusion matrices for all fitted models on the test set."}
print(results_fast$plots$cm)
```

## SHAP consensus using Spearman direction

```{r plot_ps_shap_spearman, fig.height=8, fig.cap="Top biomarkers ranked by global SHAP consensus using Spearman direction."}
print(results_fast$plots$shap_spearman)
```

## SHAP consensus using mean SHAP direction

```{r plot_ps_shap_mean, fig.height=8, fig.cap="Top biomarkers ranked by global SHAP consensus using mean SHAP direction."}
print(results_fast$plots$shap_mean)
```

## Global SHAP beeswarm

```{r plot_ps_shap_beeswarm, fig.height=8, fig.cap="Global SHAP beeswarm plot across the selected models."}
print(results_fast$plots$shap_beeswarm)
```

## Taxa prevalence

```{r plot_ps_prevalence, fig.height=7, fig.cap="Prevalence of top SHAP features across outcome groups."}
print(results_fast$plots$shap_prevalence)
```

## Integrated biomarker interpretation using Spearman direction

```{r plot_ps_integrated_spearman, fig.height=9, fig.width=12, fig.cap="Integrated biomarker interpretation using SHAP Spearman direction and prevalence."}
print(results_fast$plots$biomarker_integrated_spearman)
```

## Integrated biomarker interpretation using mean SHAP direction

```{r plot_ps_integrated_mean_shap, fig.height=9, fig.width=12, fig.cap="Integrated biomarker interpretation using mean SHAP direction and prevalence."}
print(results_fast$plots$biomarker_integrated_mean)
```

## Permutation importance heatmap

```{r plot_ps_perm_heat, fig.height=8, fig.cap="Permutation importance heatmap across models."}
print(results_fast$plots$perm_heat)
```

# Global importance tables

RUMBLE returns two global SHAP summaries. The Spearman-based direction is the
recommended option for biological interpretation because it relates feature
abundance to SHAP contribution. The mean SHAP direction instead reflects the
average direction in which the model pushed predictions when that feature was
used and should therefore be interpreted as an algorithmic tendency rather than
as a direct biological effect.

```{r importance_table_ps_spearman}
knitr::kable(
    head(results_fast$importance$global_importance$spearman, 10),
    digits = 4,
    caption = "Top 10 biomarkers by global SHAP consensus using Spearman direction."
)
```

```{r importance_table_ps_mean}
knitr::kable(
    head(results_fast$importance$global_importance$mean_shap, 10),
    digits = 4,
    caption = "Top 10 biomarkers by global SHAP consensus using mean SHAP direction."
)
```

# Model-specific beeswarm plots

By default, the exploratory run keeps `shap_model = "all"`, which means that
RUMBLE also stores isolated SHAP outputs for each model. The following chunk
iterates through all available model-specific beeswarm plots.

```{r model_specific_beeswarms, results='asis', fig.height=8}
model_names <- names(results_fast$plots$model_specific$shap_beeswarm)

for (model_name in model_names) {
    cat("### ", model_name, " beeswarm\n\n", sep = "")
    print(results_fast$plots$model_specific$shap_beeswarm[[model_name]])
    cat("\n")
}
```

# Model-specific integrated biomarker interpretation using Spearman

The next section displays the integrated Spearman-based biomarker interpretation
for each individual model, again without applying any metric cut-offs.

```{r model_specific_integrated_spearman, results='asis', fig.height=9, fig.width=12}
model_names <- names(results_fast$plots$model_specific$biomarker_integrated_spearman)

for (model_name in model_names) {
    cat("### ", model_name, " integrated biomarker interpretation (Spearman)\n\n", sep = "")
    print(results_fast$plots$model_specific$biomarker_integrated_spearman[[model_name]])
    cat("\n")
}
```

# Final remarks

This vignette intentionally keeps the workflow minimal. It shows a single fast
exploratory execution of `RUMBLE()` and then inspects the full set of core
outputs, including performance summaries, global SHAP consensus, global
integrated plots, and the model-specific beeswarm and Spearman-based integrated
biomarker views.

Because no `metric_cutoffs` are supplied, every trained model that reaches the
interpretability stage is displayed in the model-specific sections above.
