---
title: "Using RUMBLE for Microbiome Biomarker Discovery"
author: "Lucas Fett"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Using RUMBLE for Microbiome Biomarker Discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 9,
    fig.height = 7,
    fig.align = "center",
    out.width = "100%",
    dpi = 72,
    warning = FALSE,
    message = FALSE,
    crop = NULL 
)
```
# Introduction

The **RUMBLE** (Robust Unified Microbiome Biomarker Learning Engine)
package provides a streamlined workflow for identifying robust microbial
biomarkers from high-throughput sequencing data using a consensus machine
learning approach.

The pipeline executes the following steps automatically:

1. **Preprocessing**: Taxonomic aggregation, optional removal of
   unclassified taxa, prevalence filtering, and CLR transformation.

2. **Training**: Random Forest, XGBoost, Elastic Net, and KNN via
   `tidymodels`.

3. **Tuning**: Hyperparameter optimization with stratified
   cross-validation.

4. **Evaluation**: Performance metrics (F1-Score, AUC, Accuracy,
   Balanced Accuracy) and confusion matrices.

5. **Interpretability**: SHAP values and permutation importance via
   `DALEX`.

This vignette demonstrates the full analysis pipeline using two input
formats: a `phyloseq` object and a count matrix.

# Part 1: Analysis with a phyloseq Object

## Loading Data

First, we load the package and the example dataset provided.

```{r load_phyloseq}
library(RUMBLE)
library(phyloseq)
library(dplyr)
library(ggplot2)

# Load example data
ps_path <- system.file("extdata",
    "PRJEB38465_phyloseq_com_metadados_completos.rds",
    package = "RUMBLE")

if (ps_path == "") {
    stop("Example data not found. Check package installation.")
}

ps <- readRDS(ps_path)
print(ps)
```

## Exploring Metadata

```{r explore_metadata}
meta <- as.data.frame(sample_data(ps))
cat("Target variable distribution (ses):\n")
print(table(meta$ses))
```

## Scenario A: Publication-Quality Analysis (Reference Only)

The default parameters in RUMBLE are configured for robust, publication-ready
analysis. For real-world studies,
you should run the pipeline as follows:

```{r publication_quality_code, eval=FALSE}
# This code block is for demonstration purposes and is not executed here.
results_pub <- RUMBLE(
    input             = ps,
    outcome_var       = "ses",
    class_of_interest = "Low",        
    tax_level         = "Genus",
    project_name      = "My_Study_Analysis",
    output_dir        = "results_folder",  # Saves plots/tables to disk
    n_cores           = 4                  # Use parallel processing
)
```

## Scenario B: Fast Exploratory Analysis (Executed)

For this tutorial, we will run a fast version of the pipeline with reduced
parameters to demonstrate functionality quickly. We reduce the grid size and
repetitions to minimize computation time.

```{r run_exploratory, results='hide'}
set.seed(42)

# Run pipeline with reduced parameters for speed
results_fast <- RUMBLE(
    input             = ps,
    outcome_var       = "ses",
    class_of_interest = "Low",        
    tax_level         = "Genus",
    project_name      = "Fast_Exploration",
    output_dir        = tempdir(),
    grid_size         = 10,
    shap_reps         = 10,
    xgb_trees         = 100,
    rf_trees          = 100,
    n_cores           = 4
)
```

## Results: Metrics Table

```{r metrics_table_ps}
knitr::kable(results_fast$metrics, digits = 3,
             caption = "Model performance metrics (phyloseq input).")
```

## Results: All Visualizations

### 1. Metrics Comparison

```{r plot_ps_metrics, fig.height=5, fig.cap="Barplot comparing performance metrics across models."}
print(results_fast$plots$metrics)
```

### 2. ROC Curves

```{r plot_ps_roc, fig.height=5, fig.cap="ROC curves with AUC values for each model."}
print(results_fast$plots$roc)
```

### 3. Confusion Matrices

```{r plot_ps_cm, fig.height=8, fig.cap="Confusion matrices for each model on the test set."}
print(results_fast$plots$cm)
```

### 4. SHAP Consensus (Global Biomarkers)

```{r plot_ps_shap, fig.height=8, fig.cap="Top biomarkers ranked by mean absolute SHAP across all models."}
print(results_fast$plots$shap)
```

### 5. SHAP Distribution by Model

```{r plot_ps_shap_dist, fig.height=8, fig.cap="Distribution of SHAP values for top features, broken down by model."}
print(results_fast$plots$shap_dist)
```

### 6. Permutation Importance (Barplot)

```{r plot_ps_perm, fig.height=7, fig.cap="Top features ranked by permutation importance (delta loss)."}
print(results_fast$plots$perm)
```

### 7. Permutation Importance (Heatmap)

```{r plot_ps_perm_heat, fig.height=8, fig.cap="Heatmap of permutation importance across models."}
print(results_fast$plots$perm_heat)
```

## Global Importance Table
**Column Definitions:**

* **`mean_shap`**: Represents the global importance magnitude. It is calculated as the **mean of the absolute SHAP values** ($|SHAP|$).
* **`direction`**: Indicates the stability of the effect's sign. It is the **mean of the sign** of the SHAP values (ranging from -1 for consistently negative effects to +1 for consistently positive effects).

```{r importance_table_ps}
knitr::kable(
    head(results_fast$importance$global_importance, 15),
    digits = 4,
    caption = "Top 15 biomarkers by SHAP consensus (phyloseq input)."
)
```


# Part 2: Analysis with a Count Matrix

RUMBLE also accepts a standard count matrix and a separate metadata
data.frame. We demonstrate this by extracting the data from our
`phyloseq` object.

## Extracting Matrix and Metadata

```{r extract_matrix}
# Load dependencies
library(microbiome)
library(phyloseq)

# 1. Taxonomic Aggregation (Genus Level)

if ("Genus" %in% rank_names(ps)) {
  ps_genus <- microbiome::aggregate_taxa(ps, level = "Genus")
  message("Aggregated data for Gender level via tax_table.")
} else {
  warning("Rank 'Genus' not found. Using raw data..")
  ps_genus <- ps
}

# 2. Extraction of the Count Matrix (features)
otu_mat <- as.data.frame(otu_table(ps_genus))

# Machine Learning requires samples in the rows.
if (taxa_are_rows(ps_genus)) {
    otu_mat <- as.data.frame(t(otu_mat))
}

# 3. Metadata Extraction (target)
meta_df <- as.data.frame(sample_data(ps_genus))

# Integrity Verification (Sanity Check)
stopifnot(all(rownames(otu_mat) == rownames(meta_df)))
cat("Dimensões Finais: ", nrow(otu_mat), " amostras x ", ncol(otu_mat), " features.\n")
```

## Running the Pipeline

When using a count matrix, set `tax_level = NULL` since there is no
taxonomic table for aggregation.

```{r run_matrix, results='hide'}
set.seed(42)

results_mat <- RUMBLE(
    input             = otu_mat,
    metadata          = meta_df,
    outcome_var       = "ses",
    class_of_interest = "Low",           
    tax_level         = NULL,
    project_name      = "Matrix_Analysis",
    grid_size         = 10,
    shap_reps         = 10,
    xgb_trees         = 100,
    rf_trees          = 100,
    n_cores           = 4
)
```

## Verification

We verify the results by inspecting the metrics table and the ROC curve.
The results should be identical to the phyloseq run.

```{r metrics_matrix}
knitr::kable(results_mat$metrics, digits = 3,
             caption = "Model performance metrics (Matrix Input).")
```

```{r plot_matrix_roc, fig.cap="ROC curves for matrix input."}
print(results_mat$plots$roc)
```

Other plots are available in `results_mat$plots` but are omitted here for brevity.

# Session Information

```{r session_info}
sessionInfo()
```
