---
title: "Using RUMBLE for Microbiome Biomarker Discovery"
author: "Lucas Fett"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Using RUMBLE for Microbiome Biomarker Discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 9,
    fig.height = 7,
    fig.align = "center",
    out.width = "100%",
    dpi = 72,
    warning = FALSE,
    message = FALSE,
    crop = NULL 
)
```
# Introduction

The **RUMBLE** (Robust Unified Microbiome Biomarker Learning Engine)
package provides a streamlined workflow for identifying robust microbial
biomarkers from high-throughput sequencing data using a consensus machine
learning approach.

The pipeline executes the following steps automatically:

1. **Preprocessing**: Taxonomic aggregation, optional removal of
   unclassified taxa, prevalence filtering, and CLR transformation.

2. **Training**: Random Forest, XGBoost, Elastic Net, and KNN via
   `tidymodels`.

3. **Tuning**: Hyperparameter optimization with stratified
   cross-validation.

4. **Evaluation**: Performance metrics (F1-Score, AUC, Accuracy,
   Balanced Accuracy) and confusion matrices.

5. **Interpretability**: SHAP values and permutation importance via
   `DALEX`, with options for test or train data evaluation.

This vignette demonstrates the full analysis pipeline using two input
formats: a `phyloseq` object and a count matrix.

# Part 1: Analysis with a phyloseq Object

## Loading Data

First, we load the package and the example dataset provided.

```{r load_phyloseq}
library(RUMBLE)
library(phyloseq)
library(dplyr)
library(ggplot2)

# Load example data
ps_path <- system.file("extdata",
    "PRJEB38465_phyloseq_com_metadados_completos.rds",
    package = "RUMBLE")

if (ps_path == "") {
    stop("Example data not found. Check package installation.")
}

ps <- readRDS(ps_path)
print(ps)
```

## Exploring Metadata

```{r explore_metadata}
meta <- as.data.frame(sample_data(ps))
cat("Target variable distribution (ses):\n")
print(table(meta$ses))
```

## Scenario A: Publication-Quality Analysis (Reference Only)

The default parameters in RUMBLE are configured for robust, publication-ready
analysis. For real-world studies, you should run the pipeline as follows:

```{r publication_quality_code, eval=FALSE}
# This code block is for demonstration purposes and is not executed here.
results_pub <- RUMBLE(
    input             = ps,
    outcome_var       = "ses",
    class_of_interest = "Low",        
    tax_level         = "Genus",
    project_name      = "My_Study_Analysis",
    output_dir        = "results_folder",  # Saves plots/tables to disk
    shap_data         = "test",            # Default: evaluate SHAP on test set
    n_cores           = 4                  # Use parallel processing
)
```

## Scenario B: Fast Exploratory Analysis (Executed)

For this tutorial, we will run a fast version of the pipeline with reduced
parameters to demonstrate functionality quickly. We reduce the grid size and
repetitions to minimize computation time. We explicitly define `shap_data = "test"` (the default).

```{r run_exploratory, results='hide'}
set.seed(42)

# Run pipeline with reduced parameters for speed
results_fast <- RUMBLE(
    input             = ps,
    outcome_var       = "ses",
    class_of_interest = "Low",        
    tax_level         = "Genus",
    project_name      = "Fast_Exploration",
    output_dir        = tempdir(),
    grid_size         = 10,
    shap_reps         = 2,
    xgb_trees         = 100,
    rf_trees          = 100,
    n_cores           = 4,
    shap_data         = "test" # Using test set for SHAP
)
```

## Results: Metrics Table

```{r metrics_table_ps}
knitr::kable(results_fast$metrics, digits = 3,
             caption = "Model performance metrics (phyloseq input).")
```

## Results: All Visualizations

### 1. Metrics Comparison

```{r plot_ps_metrics, fig.height=5, fig.cap="Barplot comparing performance metrics across models."}
print(results_fast$plots$metrics)
```

### 2. ROC Curves

```{r plot_ps_roc, fig.height=5, fig.cap="ROC curves with AUC values for each model."}
print(results_fast$plots$roc)
```

### 3. Confusion Matrices

```{r plot_ps_cm, fig.height=8, fig.cap="Confusion matrices for each model on the test set."}
print(results_fast$plots$cm)
```

### 4. SHAP Consensus (Spearman & Mean SHAP)

RUMBLE calculates biomarker consensus using two metrics for effect direction: Spearman correlation and Mean SHAP.

```{r plot_ps_shap_spearman, fig.height=8, fig.cap="Top biomarkers ranked by SHAP consensus (Spearman direction)."}
print(results_fast$plots$shap_spearman)
```

```{r plot_ps_shap_mean, fig.height=8, fig.cap="Top biomarkers ranked by SHAP consensus (Mean SHAP direction)."}
print(results_fast$plots$shap_mean)
```

### 5. SHAP Distribution (Beeswarm)

```{r plot_ps_shap_beeswarm, fig.height=8, fig.cap="Beeswarm plot of SHAP values showing feature distribution."}
print(results_fast$plots$shap_beeswarm)
```

### 6. Taxa Prevalence 

```{r plot_ps_prevalence, fig.height=7, fig.cap="Prevalence of top SHAP features across groups."}
print(results_fast$plots$shap_prevalence)
```

### 7. Integrated Biomarker Interpretation

RUMBLE generates the integrated plots using both Spearman and Mean SHAP directions.
```{r plot_ps_integrated_mean_shap, fig.height=9, fig.width=12, fig.cap="Combined view of SHAP importance (Mean SHAP) and group prevalence."}
print(results_fast$plots$biomarker_integrated_mean)
```

```{r plot_ps_integrated_spearman, fig.height=9, fig.width=12, fig.cap="Combined view of SHAP importance (Spearman) and group prevalence."}
print(results_fast$plots$biomarker_integrated_spearman)
```

### 8. Permutation Importance (Heatmap)

```{r plot_ps_perm_heat, fig.height=8, fig.cap="Heatmap of permutation importance across models."}
print(results_fast$plots$perm_heat)
```

## Global Importance Tables

RUMBLE outputs global importance using two distinct metrics to infer feature directionality:

* **Spearman Correlation (`direction`)**: Indicates the correlation between the original feature value and its SHAP contribution. Highly robust for monotonic relationships.
* **Mean SHAP (`direction`)**: The mean of the sign of the SHAP values. Useful to see if a feature predominantly pushes predictions positive or negative.

```{r importance_table_ps_spearman}
knitr::kable(
    head(results_fast$importance$global_importance$spearman, 10),
    digits = 4,
    caption = "Top 10 biomarkers by SHAP consensus (Spearman)."
)
```

```{r importance_table_ps_mean}
knitr::kable(
    head(results_fast$importance$global_importance$mean_shap, 10),
    digits = 4,
    caption = "Top 10 biomarkers by SHAP consensus (Mean SHAP)."
)
```


# Part 2: Analysis with a Count Matrix

RUMBLE also accepts a standard count matrix and a separate metadata
data.frame. We demonstrate this by extracting the data from our
`phyloseq` object. In this scenario, we will explicitly set `shap_data = "train"` to demonstrate interpretation on the training data.

## Extracting Matrix and Metadata

```{r extract_matrix}
# Load dependencies
library(microbiome)
library(phyloseq)

# 1. Taxonomic Aggregation (Genus Level)

if ("Genus" %in% rank_names(ps)) {
  ps_genus <- microbiome::aggregate_taxa(ps, level = "Genus")
  message("Aggregated data for Gender level via tax_table.")
} else {
  warning("Rank 'Genus' not found. Using raw data..")
  ps_genus <- ps
}

# 2. Extraction of the Count Matrix (features)
otu_mat <- as.data.frame(otu_table(ps_genus))

# Machine Learning requires samples in the rows.
if (taxa_are_rows(ps_genus)) {
    otu_mat <- as.data.frame(t(otu_mat))
}

# 3. Metadata Extraction (target)
meta_df <- as.data.frame(sample_data(ps_genus))

# Integrity Verification (Sanity Check)
stopifnot(all(rownames(otu_mat) == rownames(meta_df)))
cat("Dimensões Finais: ", nrow(otu_mat), " amostras x ", ncol(otu_mat), " features.\n")
```

## Running the Pipeline (Using Train Data for SHAP)

When using a count matrix, set `tax_level = NULL` since there is no
taxonomic table for aggregation. Here, we switch `shap_data = "train"`.

```{r run_matrix, results='hide'}
set.seed(42)

results_mat <- RUMBLE(
    input             = otu_mat,
    metadata          = meta_df,
    outcome_var       = "ses",
    class_of_interest = "Low",            
    tax_level         = NULL,
    project_name      = "Matrix_Analysis",
    grid_size         = 10,
    shap_reps         = 2,
    xgb_trees         = 100,
    rf_trees          = 100,
    n_cores           = 4,
    shap_data         = "train" # Switch to training set for interpretability
)
```

## Verification

We verify the results by inspecting the metrics table and the ROC curve. The performance metrics remain evaluated on the test set, but interpretability plots (like SHAP) for this specific run are derived from the training data.

```{r metrics_matrix}
knitr::kable(results_mat$metrics, digits = 3,
             caption = "Model performance metrics (Matrix Input).")
```

```{r plot_matrix_roc, fig.cap="ROC curves for matrix input."}
print(results_mat$plots$roc)
```
## Interpretability on Training Data

Because we explicitly set `shap_data = "train"` in this second run, the interpretability plots now reflect the feature importance derived directly from the training set. This can be particularly useful to understand the baseline patterns the model learned before generalizing, or to check for potential overfitting when compared to test set explanations.

```{r plot_matrix_shap_spearman, fig.height=8, fig.cap="Top biomarkers ranked by SHAP consensus on Training Data (Spearman)."}
print(results_mat$plots$shap_spearman)
```

```{r plot_matrix_shap_mean, fig.height=8, fig.cap="Top biomarkers ranked by SHAP consensus on Training Data (Spearman)."}
print(results_mat$plots$shap_mean)
```
# Integrated Biomarker Interpretation (Training Data)
Just like the test set evaluation, we can view the integrated interpretation combining the SHAP consensus and the ecological prevalence directly on the training data.
```{r integrated_spearman, fig.height=9, fig.width=12, fig.cap="Combined view of SHAP importance (Mean SHAP) and group prevalence."}
print(results_mat$plots$biomarker_integrated_spearman)
```
```{r integrated_meanplot_ps_integrated_spearman, fig.height=9, fig.width=12, fig.cap="Combined view of SHAP importance (Spearman) and group prevalence."}
print(results_mat$plots$biomarker_integrated_mean)
```

```{r matrix_bee}
print(results_mat$plots$shap_beeswarm)
```
Other plots are available in `results_mat$plots` but are omitted here for brevity.

# Session Information

```{r session_info}
sessionInfo()
```
